### 2015-02-26 Project_Update/Update.tex
Project update-- missing the citation for the red book Rachel lent us. I'll add it in tomorrow morning and print.

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\begin{document}
\begin{titlepage}
\title{CPSC 504 Project Update \vspace{4 cm}}
\date{\vspace{2 cm} February 26, 2015}

\author{
 \makebox[1.0\linewidth]{Laura Cang}\\cang@cs.ubc.ca\\
  \and \makebox[1.0\linewidth]{Kailang Jiang}\\jiangkl@cs.ubc.ca\\
  \and \makebox[1.0\linewidth]{Jessica Wong}\\jhmwong@cs.ubc.ca\\ 
}
\maketitle
\thispagestyle{empty}
\end{titlepage}
\newpage
\section{What has been done so far}
We have decided to focus on a portion of Arni's thesis \cite{arniThesis} as our project. Specifically, we have chosen to implement something that will help solve Scenario 1, a situation where there is a user who has pulled some data from the database and in the process has obtained a local copy of a table along with a copy of the schema mapping that relates how the columns in the local table relate to the columns in the table that is stored in the database. The user has then modified the data somehow by making some changes; the type of changes a user can make are limited to (i.e., adding a row, deleting a row, adding a column, deleting a column, updating a column name, or updating a value in a cell). As the user makes various changes, his/her actions will be tracked in the background in a transformation script. This transformation script will be written in some declarative language that we have not exactly determined yet (it will require more research into ETL languages). When the user wishes to push the local data and its various changes back to the database, he/she will be able to do so. On the back end, the transformation script will be executed as a set of instructions to the database in order to save the changes remotely; if there are schema changes to be made, these schema mapping changes will also be done at this time. The bulk of our project is going to focus on how this transformation script is defined along with how it will work with the schema mappings and what kind of output script will result from a user's push command.

In our project, we wish to implement an interface that will allow the user to view the data he/she has pulled as well as conduct the various actions mentioned above. Our measure of success will be whether or not we can successfully complete Scenario 1 using our implementation. We will be focusing on trying to building an interface that can successfully transform data changes created by the user to a transformation script and using both the transformation script and the schema mapping to push correct changes to the datastore. We are not focusing on having a wondrously usable GUI nor will we be focusing on creating a script to pull/push data from a database to our interface or vice versa.

We plan to test with a database of 5 tables that have about 500 tuples altogether. This is anticipated to be one of the datasets that the CPSC 304 classes use for their project.

\section{Future Work}
If there is time, we will consider the situation where the user can pull data from more than one table, and perform data transformations on it.

\section{What there is left to do}
\begin{itemize}
	\item{Define the type of transformations allowed in our scenario}
	\item{Design decisions about the transformation language}
	\item{Design how to maintain the mapping}
	\item{Decide how to handle various actions}
	\item{Decide what kind of output do we want when the user pushes}
	\item{Decide how to push local updates to the remote server}
	\item{Design and implement the interface}
\end{itemize}

\section{Roadblocks}
\begin{itemize}
	\item{Need data to test with}
	\item{Choose or define a transformation language}
\end{itemize}

\section{Related Work}
One of the largest problems that we foresee is creating a transformation language to track all the possible data changes as well as how to propagate those changes correctly from one source to another. Hence, most of our related work focuses on systems that are similar or areas of work where we feel we can draw inspiration from how they tackled a similar problem to the one we are facing.

\subsection{Data Sharing Systems}
Data sharing situations are not new; there have been many collaborative data sharing systems (CDSS) \cite{ives2008orchestra, ives2005orchestra, karvounarakis2013collaborative}and data transformation language \cite{kandel2011wrangler, lakshmanan2001schemasql} created to try and address this issue. From CDSS systems, we believe we can examine how it handles schema mapping change propagations to inform our own design as this is exactly the same problem we are facing.

\subsection{Transformation Languages}
As mentioned before, a large part of our project involves tracking the changes a user has made to the data. Transformation languages capture this information but we currently have no idea how to structure this transformation language. We are hoping to look at other transformation languages \cite{kandel2011wrangler, lakshmanan2001schemasql} and research related to transformation languages \cite{abiteboul1999tools, kandel2011wrangler, lakshmanan2001schemasql, raman2001potter} to see if there is a suitable one we can work off of. If we cannot find one suitable for our needs, we can use these languages to identify what principles are most important to have in a transformation language and implement them in our own version. We are leaning towards the first option rather than the second.

\subsection{Extraction, Transformation, and Loading (ETL)}
Extraction, transformation, and loading (ETL) \cite{vassiliadis2009extraction}[and the red book] is the process of taking data from one source, changing it, and then putting it into a data warehouse. Although this is not exactly what we are doing, we can still examine how it is done to figure out how to best track the changes made to the data. One of the biggest problems we can foresee for our project is how we can track the various changes made to our data once the user has pulled a copy of the data to his/her interface. We believe that we can use some of the ETL principles to guide how we can best do this process.

\subsection{View Maintenance}
Currently we have chosen to only allow data from a single table be viewed in our interface. However, given that this project could be extended in the future to include viewing data that is from multiple tables, we are looking ahead at literature in the view maintenance area to try and predict what challenges our interface might face \cite{ agrawal1997efficient, agrawal2009asynchronous, gupta1999materialized, zhou2007lazy}. We also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.

\section{Timeline}
\begin{tabular}{| l | l |}
\hline
Feb 22 - Feb 28 & Start literature review on data sharing systems and transformation languages\\
\hline
Mar 1 - Mar 7 & Literature review \\
& NOTES: Jessica and Laura both gone for most of this week \\
\hline
Mar 8 - Mar 14 & Finish literature review on Monday\\
& Start implementation\\
\hline
Mar 15 - Mar 21 & Finish implementation\\
& Start testing\\
\hline
Mar 22 - Apr 4 & Finish testing\\
& Start writeup and prepare for presentatino\\
\hline
Apr 5 - Apr 11 & Handin paper\\
&\\
\hline
\end{tabular}
\newpage
\bibliographystyle{abbrv}
\bibliography{504_Update}

\end{document}

### 2015-02-28 Project_Update/Update.tex
Phase II write-up.

@@
 \subsection{Extraction, Transformation, and Loading (ETL)}
-Extraction, transformation, and loading (ETL) \cite{vassiliadis2009extraction}[and the red book] is the process of taking data from one source, changing it, and then putting it into a data warehouse.
+Extraction, transformation, and loading (ETL) \cite{vassiliadis2009extraction} is the process of taking data from one source, changing it, and then putting it into a data warehouse.
 Although this is not exactly what we are doing, we can still examine how it is done to figure out how to best track the changes made to the data.

### 2015-04-06 src/gitFrame.java
New GUI and test data

@@ -1,0 +1,337 @@
+/*
+ * To change this license header, choose License Headers in Project Properties.
+ * To change this template file, choose Tools | Templates
+ * and open the template in the editor.
+ */
+
+import java.io.BufferedReader;
+import java.io.FileReader;
+import java.io.IOException;
+import java.io.PrintWriter;
+import java.util.logging.Level;
+import java.util.logging.Logger;
+import javax.swing.table.DefaultTableModel;
+import javax.swing.table.TableColumn;
+import javax.swing.table.TableColumnModel;
+
+/**
+ *
+ * @author jkl
+ */
+public class gitFrame extends javax.swing.JFrame {
+
+    /**
+     * Creates new form gitFrame
+     */
+    String loadPath= "data.csv";
+    String logPath= "log.script";
+    BufferedReader br;
+    PrintWriter pw;
+    
+    public gitFrame() {
+        initComponents();
+    }
+    
+    protected void finalize(){
+        try {
+            super.finalize();
+            pw.close();
+        } catch (Throwable ex) {
+            Logger.getLogger(gitFrame.class.getName()).log(Level.SEVERE, null, ex);
+        }
+    };
+
+    /**
+     * This method is called from within the constructor to initialize the form.
+     * WARNING: Do NOT modify this code. The content of this method is always
+     * regenerated by the Form Editor.
+     */
+    @SuppressWarnings("unchecked")
+    // <editor-fold defaultstate="collapsed" desc="Generated Code">//GEN-BEGIN:initComponents
+    private void initComponents() {
+
+        addRowBtn = new javax.swing.JButton();
+        addColBtn = new javax.swing.JButton();
+        delRowBtn = new javax.swing.JButton();
+        delColBtn = new javax.swing.JButton();
+        loadBtn = new javax.swing.JButton();
+        pushBtn = new javax.swing.JButton();
+        jScrollPane1 = new javax.swing.JScrollPane();
+        myTable = new javax.swing.JTable();
+        clearBtn = new javax.swing.JButton();
+
+        setDefaultCloseOperation(javax.swing.WindowConstants.EXIT_ON_CLOSE);
+        setTitle("CPSC 504 Project");
+
+        addRowBtn.setText("Add row");
+        addRowBtn.setHorizontalAlignment(javax.swing.SwingConstants.LEFT);
+        addRowBtn.addActionListener(new java.awt.event.ActionListener() {
[…]

### 2015-04-06 src/gitFrame.java
Minor modification to the GUI

@@ -4,6 +4,7 @@
  * and open the template in the editor.
  */
 
+
 import java.io.BufferedReader;
 import java.io.FileReader;
 import java.io.IOException;
@@ -59,9 +60,12 @@ public class gitFrame extends javax.swing.JFrame {
         jScrollPane1 = new javax.swing.JScrollPane();
         myTable = new javax.swing.JTable();
         clearBtn = new javax.swing.JButton();
+        colNameTextField = new javax.swing.JTextField();
+        updColBtn = new javax.swing.JButton();
 
         setDefaultCloseOperation(javax.swing.WindowConstants.EXIT_ON_CLOSE);
         setTitle("CPSC 504 Project");
+        setResizable(false);
 
         addRowBtn.setText("Add row");
         addRowBtn.setHorizontalAlignment(javax.swing.SwingConstants.LEFT);
@@ -117,12 +121,24 @@ public class gitFrame extends javax.swing.JFrame {
         ));
         myTable.setToolTipText("");
         myTable.setColumnSelectionAllowed(true);
+        myTable.getTableHeader().setReorderingAllowed(false);
         jScrollPane1.setViewportView(myTable);
 
         clearBtn.setText("Clear all");
         clearBtn.addActionListener(new java.awt.event.ActionListener() {
             public void actionPerformed(java.awt.event.ActionEvent evt) {
                 clearBtnActionPerformed(evt);
+            }
+        });
+
+        colNameTextField.setHorizontalAlignment(javax.swing.JTextField.LEFT);
+        colNameTextField.setText("Column Name");
+
+        updColBtn.setText("Update Column");
+        updColBtn.setHorizontalAlignment(javax.swing.SwingConstants.LEFT);
+        updColBtn.addActionListener(new java.awt.event.ActionListener() {
+            public void actionPerformed(java.awt.event.ActionEvent evt) {
+                updColBtnActionPerformed(evt);
             }
         });
 
@@ -136,12 +152,16 @@ public class gitFrame extends javax.swing.JFrame {
                     .addComponent(jScrollPane1, javax.swing.GroupLayout.PREFERRED_SIZE, 715, javax.swing.GroupLayout.PREFERRED_SIZE)
                     .addGroup(layout.createSequentialGroup()
                         .addGroup(layout.createParallelGroup(javax.swing.GroupLayout.Alignment.LEADING, false)
[…]

### 2015-04-07 src/gitFrame.java
Able to log script

@@ -6,14 +6,18 @@
 
 
 import java.io.BufferedReader;
+import java.io.FileNotFoundException;
+import java.io.FileOutputStream;
 import java.io.FileReader;
 import java.io.IOException;
 import java.io.PrintWriter;
+import java.io.UnsupportedEncodingException;
+import java.util.ArrayList;
 import java.util.logging.Level;
 import java.util.logging.Logger;
+import javax.swing.event.TableModelEvent;
+import javax.swing.event.TableModelListener;
 import javax.swing.table.DefaultTableModel;
-import javax.swing.table.TableColumn;
-import javax.swing.table.TableColumnModel;
 
 /**
  *
@@ -26,10 +30,24 @@ public class gitFrame extends javax.swing.JFrame {
      */
     String loadPath= "data.csv";
     String logPath= "log.script";
+    String pushPath= "data_new.csv";
     BufferedReader br;
     PrintWriter pw;
+    BufferedReader logbr;
+    
+    boolean sync= true;
+    boolean[] syncs= new boolean[100];
+    
+    ArrayList<String> headers;
+    ArrayList<ArrayList<String>> dataset;
+    //int nTempCol=0;
     
     public gitFrame() {
+        try {
+            pw = new PrintWriter(new FileOutputStream(logPath), true);
+        } catch (FileNotFoundException ex) {
+            Logger.getLogger(gitFrame.class.getName()).log(Level.SEVERE, null, ex);
+        }        
         initComponents();
     }
     
@@ -62,6 +80,7 @@ public class gitFrame extends javax.swing.JFrame {
         clearBtn = new javax.swing.JButton();
         colNameTextField = new javax.swing.JTextField();
         updColBtn = new javax.swing.JButton();
+        syncCheckBox = new javax.swing.JCheckBox();
 
         setDefaultCloseOperation(javax.swing.WindowConstants.EXIT_ON_CLOSE);
         setTitle("CPSC 504 Project");
@@ -110,6 +129,11 @@ public class gitFrame extends javax.swing.JFrame {
 
         pushBtn.setFont(new java.awt.Font("Lucida Grande", 1, 13)); // NOI18N
         pushBtn.setText("Push");
+        pushBtn.addActionListener(new java.awt.event.ActionListener() {
+            public void actionPerformed(java.awt.event.ActionEvent evt) {
+                pushBtnActionPerformed(evt);
+            }
+        });
 
[…]

### 2015-04-07 src/gitFrame.java
Push enabled

@@ -11,7 +11,6 @@ import java.io.FileOutputStream;
 import java.io.FileReader;
 import java.io.IOException;
 import java.io.PrintWriter;
-import java.io.UnsupportedEncodingException;
 import java.util.ArrayList;
 import java.util.logging.Level;
 import java.util.logging.Logger;
@@ -28,9 +27,9 @@ public class gitFrame extends javax.swing.JFrame {
     /**
      * Creates new form gitFrame
      */
-    String loadPath= "data.csv";
+    String dataPath= "data.csv";
     String logPath= "log.script";
-    String pushPath= "data_new.csv";
+    String initDataPath= "data_init.csv";
     BufferedReader br;
     PrintWriter pw;
     BufferedReader logbr;
@@ -51,6 +50,7 @@ public class gitFrame extends javax.swing.JFrame {
         initComponents();
     }
     
+    @Override
     protected void finalize(){
         try {
             super.finalize();
@@ -77,13 +77,13 @@ public class gitFrame extends javax.swing.JFrame {
         pushBtn = new javax.swing.JButton();
         jScrollPane1 = new javax.swing.JScrollPane();
         myTable = new javax.swing.JTable();
-        clearBtn = new javax.swing.JButton();
+        resetBtn = new javax.swing.JButton();
         colNameTextField = new javax.swing.JTextField();
         updColBtn = new javax.swing.JButton();
         syncCheckBox = new javax.swing.JCheckBox();
 
         setDefaultCloseOperation(javax.swing.WindowConstants.EXIT_ON_CLOSE);
-        setTitle("CPSC 504 Project");
+        setTitle("CPSC 504 Project - DBGit");
         setResizable(false);
 
         addRowBtn.setText("Add row");
@@ -159,10 +159,10 @@ public class gitFrame extends javax.swing.JFrame {
             }
         });
 
-        clearBtn.setText("Clear all");
-        clearBtn.addActionListener(new java.awt.event.ActionListener() {
+        resetBtn.setText("Reset");
+        resetBtn.addActionListener(new java.awt.event.ActionListener() {
             public void actionPerformed(java.awt.event.ActionEvent evt) {
-                clearBtnActionPerformed(evt);
+                resetBtnActionPerformed(evt);
             }
         });
 
@@ -208,7 +208,7 @@ public class gitFrame extends javax.swing.JFrame {
                         .addGap(18, 18, 18)
[…]

### 2015-04-08 Final_Paper/Paper.tex
Draft of the Future Work section.

\documentclass[12pt]{article}
\begin{document}
\begin{titlepage}
\title{\textbf{CPSC 504 Project\\\vspace{3 cm}\Huge{An investigation into an application to manage data transformations and updates}\vspace{2 cm}}}
\date{\vspace{2 cm} April 14, 2015}

\author{
 \makebox[1.0\linewidth]{Laura Cang}\\cang@cs.ubc.ca\\
  \and \makebox[1.0\linewidth]{Kailang Jiang}\\jiangkl@cs.ubc.ca\\
  \and \makebox[1.0\linewidth]{Jessica Wong}\\jhmwong@cs.ubc.ca\\ 
}
\maketitle
\thispagestyle{empty}
\end{titlepage}

\newpage
\section{Introduction}
The basis

\section{Motivation}
In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables, it is important to figure out how this relates to the real world.

\subsection{Example Scenario}
In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application. This was a movie dataset that we decided to take three columns out of to use

\section{Solution}
\subsection{Transformation Language}
\subsection{View Maintenance}
\subsection{Proof of Concept}

\section{Related Work}

\section{Future Work}
Future extensions of this work can look into testing the transformation language on other data formats like XML to see if it can effectively capture the changes that happen in an XML document. We can also look into how to handle situations where there are multiple concurrent users drawing upon the same data source and what policies should be adopted into order to achieve the best balance between performance and data consistency. Updating the data source too often could cause performance overhead from locking up the data source while updating too infrequently could cause inconsistent data to be propagated to many different users. The amount of effort and operational overhead involved with many people making the same changes would be undesirable.

Another extension of the project could be investigating how the data transformations could be integrated into data provenance to help with data conflict resolutions \cite{arniThesis}. As the project currently stands, the remote application can push any type of data to the data source. If the data being pushed to the data source violates any table constraints (e.g., if a user tries to push a non-unique value into a column that has specified that it only takes unique values), the database administrator has to manually resolve that violation. It is possible to investigate whether or not provenance information can be leveraged to help make these decisions thus lowering the workload on the database administrator. A possible example of how to use the provenance information could be using the previous history of the sorts of changes have been accepted for that column or table to determine the likelihood of the current data violation being accepted. Provenance could also be used to determine the likelihood of correctness by the app or user who made the change; as information about who has changed values can be tracked, it is not inconceivable that the statistics of who has made the best or the most correct/acceptable changes can be used in some way.

\bibliographystyle{abbrv}
\bibliography{final_paper_biblography}

\end{document}

### 2015-04-08 Final_Paper/Paper.tex
Draft of Introduction.

@@
 \section{Solution}
+In this project, we focus on three different ways a user can use the application to fetch data.
+
+\noindent
+
+\section{Types of Changes Supported}
 \subsection{Transformation Language}
@@
 \subsection{Proof of Concept}
+In the proof of concept, we had an extra step that was not included in \cite{arniThesis}.
+We decided to allow for the user to choose which changes to sync to the remote data source.
+\textcolor{red}{add screen shots!!
+Kailang can write about all the cool things she did here!!
+:)}
 

### 2015-04-08 Final_Paper/Paper.tex
This is the actual draft of the Introduction. =D

@@
 \documentclass[12pt]{article}
+\usepackage{color}
 \begin{document}
@@
 \section{Introduction}
-The basis
+This project is built on top of the theoretical framework done in the M.Sc thesis of Arni Thrastarson \cite{arniThesis} and focuses on the first of the four scenarios detailed in the work.
+The first scenario is about trying to investigate how changes made by a data application can be captured and propagated to a remote data source.
+The main workflow of this scenario would be a remote (possibly third party) application pulling the data from a data source to use locally.
+The application will receive a local copy of the data along with a schema mapping to identify which tables the columns in the local copy of the data came from.
+A user using the application may notice inconsistencies with the data and go to correct it.
+The changes made by this user are logged in a transformation script and when the time comes to update the data source, the transformation script is sent to the data store and the changes detailed in the script are applied.
+
+\noindent Our primary contributions are:
+\begin{enumerate}
+	\item{Using logical programming concepts to define a data transformation language that captures changes made to a relational data model}
+	\item{Design decisions about how the relational model should react and handle schema changes}
+	\item{A preliminary investigation into view maintenance policies for the remote datastore}
+	\item{Further detailing the scenario in which an application views a subset of columns from one or more tables in a remote dataset}
+	\item{Deciding which issues to assign to the server side and which to the client side}
+\end{enumerate}
 

### 2015-04-08 Final_Paper/Paper.tex
Draft of the other things I have.

@@
 \section{Introduction}
+\label{sec:introduction}
 This project is built on top of the theoretical framework done in the M.Sc thesis of Arni Thrastarson \cite{arniThesis} and focuses on the first of the four scenarios detailed in the work.
@@
 The changes made by this user are logged in a transformation script and when the time comes to update the data source, the transformation script is sent to the data store and the changes detailed in the script are applied.
+
+In this project, we focus on three different ways a user can use the application to fetch data.
+\begin{enumerate}
+	\item {When the user chooses to view all the columns in one table}
+	\item {When the user chooses to view a subset of the columns in one table}
+	\item {When the user chooses to view a subset of the columns in multiple tables}
+\end{enumerate}
 
@@
 In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables, it is important to figure out how this relates to the real world.
+The example in \cite{arniThesis} talks about a sports reporter who retrieves select information about basketball players to do an article on.
+In this case, the reporter only takes out columns regarding a player's name, his high school, and the first team he played on after entering the NBA.
+Compared to all the information that is stored about NBA players on a main database (e.g., salary, weight, height, number of transfers, agent name, etc.), the reporter has only taken a very small subset of columns.
+Another example could be a school administrator who wants to decide which students to give awards to purely based on academic merit.
+He/she could go to the main database storing all student information (e.g., the database could have information regarding a student's name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) and only choose to view the student number and GPA column so that he/she will not be biased towards any names or gender.
 
-\subsection{Example Scenario}
+An example where a user might need to view a subset of columns in multiple tables would be if the user is trying to find detailed information about proteins found in different microbiological organisms.
+The microbiological organisms and the various proteins it produces could be stored in one table while detailed information about each protein is stored in another.
+If a user wanted to see certain characteristics of a protein found in specific organisms, he/she would have use a subset of columns from both the microbiological organism table along with the protein time to get all the relevant information.
+
+\subsection{Motivation used in our project}
 In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application.
-This was a movie dataset that we decided to take three columns out of to use
+This was a movie dataset that we decided to take three columns out of to use \textcolor{red}{put stuff here because i don't actually know anything about the dataset}
 
 \section{Solution}
-In this project, we focus on three different ways a user can use the application to fetch data.
+The basic workflow of the scenario described in Section \ref{sec:introduction} will remain the same.
+However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformations increase along with the amount of information that must be tracked (see Section \ref{sec:transformation_language}).
 
-\noindent
+\subsection{Types of Changes Supported}
+Based on the various classes of changes mentioned in Wrangler \cite{kandel2011wrangler}, we decided to focus on two of them: map and reshape.
+Map changes deal with "[transforming] one input data row to zero, one, or multiple output rows" \cite{kandel2011wrangler}.
+This essentially means that the row changes that are supported in our application (e.g., add row, delete row, edit value(s) in row(s)) all fall within this class of changes.
+Reshape changes deal with schema changes.
+Any change to the columns (e.g., add/remove/rename columns) will fall into this category.
+The scope of this project does not deal with changing the data type of a column.
 
-\section{Types of Changes Supported}
 \subsection{Transformation Language}
+\label{sec:transformation_language}
+
 \subsection{View Maintenance}
+View maintenance \textcolor{red}{cite stuff here}
+
 \subsection{Proof of Concept}
-In the proof of concept, we had an extra step that was not included in \cite{arniThesis}.
-We decided to allow for the user to choose which changes to sync to the remote data source.
+After designing the transformation language, we decided to implement the scenario workflow in order to show that our transformation language and design decisions could work to update a data source.
+The only extra step that was not included in \cite{arniThesis} was to allow for the user to choose which changes to sync to the remote data source.
 \textcolor{red}{add screen shots!!

### 2015-04-08 Final_Paper/Paper.tex
Nothing substantial.
stuff

@@
 
-In this project, we focus on three different ways a user can use the application to fetch data.
+In this project, we focus on three different ways a user can use the application to fetch data. For the rest of this paper, we shall call each way of data fetching a sub-scenario. 
 \begin{enumerate}
@@
 \label{sec:transformation_language}
+As we designed the transformation language, we realized it did not make sense to work off the simplest of our three sub-scenarios as the transformation language did not adapt well when the complexity of the sub-scenario was increased.
+Hence, we decided to work off the most complicated of our three sub-scenarios, the situation where an application has pulled a subset of columns from more than one table, and design our transformation language from that point onwards.
+
+\textbf{\textcolor{green}{more here}}
 
@@
 \section{Related Work}
+\textcolor{red}{need stuff here can probably reuse the stuff from the project update but will need more + editing}
+\textbf{\textcolor{green}{Stuff in this section is ripped from the project update}}
+One of the largest problems that we foresee is creating a transformation language to track all the possible data changes as well as how to propagate those changes correctly from one source to another.
+Hence, most of our related work focuses on systems that are similar or areas of work where we feel we can draw inspiration from how they tackled a similar problem to the one we are facing.
+
+\subsection{Data Sharing Systems}
+Data sharing situations are not new; there have been many collaborative data sharing systems (CDSS) \cite{ives2008orchestra, ives2005orchestra, karvounarakis2013collaborative}and data transformation language \cite{kandel2011wrangler, lakshmanan2001schemasql} created to try and address this issue.
+From CDSS systems, we believe we can examine how it handles schema mapping change propagations to inform our own design as this is exactly the same problem we are facing.
+
+\subsection{Transformation Languages}
+There is no lack of data transformation tools
+
+\subsection{View Maintenance}
+Currently we have chosen to only allow data from a single table be viewed in our interface.
+However, given that this project could be extended in the future to include viewing data that is from multiple tables, we are looking ahead at literature in the view maintenance area to try and predict what challenges our interface might face \cite{ agrawal1997efficient, agrawal2009asynchronous, gupta1999materialized, zhou2007lazy}.
+We also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.
 

### 2015-04-13 Final_Paper/Paper.tex
Paper now.

@@
 \usepackage{color}
+\usepackage{tabularx}
+
 \begin{document}
@@
 This project is built on top of the theoretical framework done in the M.Sc thesis of Arni Thrastarson \cite{arniThesis} and focuses on the first of the four scenarios detailed in the work.
-The first scenario is about trying to investigate how changes made by a data application can be captured and propagated to a remote data source.
-The main workflow of this scenario would be a remote (possibly third party) application pulling the data from a data source to use locally.
-The application will receive a local copy of the data along with a schema mapping to identify which tables the columns in the local copy of the data came from.
-A user using the application may notice inconsistencies with the data and go to correct it.
-The changes made by this user are logged in a transformation script and when the time comes to update the data source, the transformation script is sent to the data store and the changes detailed in the script are applied.
+The first scenario investigates how changes made by a data application can be captured and propagated to a remote data source.
+The main workflow likely involves a (possibly third party) application pulling the data from a remote data source to use locally.
+The application creates a local copy of the data along with a schema mapping to identify the source table(s) of the data attributes.
+A user using the application may notice inconsistencies with the data and correct them.
+Any changes made are logged in a transformation script and when the time comes to update the data source, the transformation script is sent to the data store where the changes detailed in the script are applied.
 
@@
 \begin{enumerate}
-	\item{Using logical programming concepts to define a data transformation language that captures changes made to a relational data model}
+	\item{Defined a data transformation language using logical programming concepts to capture changes made to a relational data model}
 	\item{Design decisions about how the relational model should react and handle schema changes}
-	\item{A preliminary investigation into view maintenance policies for the remote datastore}
+	\item{Preliminary investigation into view maintenance policies for the remote datastore}
 	\item{Further detailing the scenario in which an application views a subset of columns from one or more tables in a remote dataset}
-	\item{Deciding which issues to assign to the server side and which to the client side}
+	\item{Decided which issues to assign to the server side and which to the client side}
 \end{enumerate}
@@
 
-\subsection{Motivation used in our project}
+\subsection{Sample data used in our project}
 In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application.
-This was a movie dataset that we decided to take three columns out of to use \textcolor{red}{put stuff here because i don't actually know anything about the dataset}
+This was a movie star dataset that had the columns ``Star Id'', ``Name'', and ``Gender''.
+After supplementing some additional data to the dataset, namely the columns ``Movie Title'', ``Production Studio'', and ``Critic Rating'', we used this dataset.
+We also added in a second table detailing user information with the columns ``Movie Title'' which is a foreign key referencing the ``Movie Title'' column in the Movie Information table, ``Website that user information was obtained from'', ``User Rating'', and ``Number of Likes''.
 
@@
 However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformations increase along with the amount of information that must be tracked (see Section \ref{sec:transformation_language}).
+In this project, we have decided to support the following user changes: adding a column, removing a column, editing a column name, changing a table cell's current value, removing a table cell's value to make it null, adding a row, and removing a row.
+Any changes that cause a violation in schema definitions (e.g., removing a row's primary key) will be resolved on the server side by the database administrator.
+This design decision was due to not wanting to restrict the types of users that would be able to use an app to find data to examine.
+As we have to assume that there could be users unfamiliar with schema/storage restrictions, we decided to ensure that schema violations are resolved by someone with the technical understanding of what error has occurred.
 
@@
 Hence, we decided to work off the most complicated of our three sub-scenarios, the situation where an application has pulled a subset of columns from more than one table, and design our transformation language from that point onwards.
+We then retroactively checked to ensure that the transformations worked for the simpler sub-scenarios.
 
-\textbf{\textcolor{green}{more here}}
+When designing the data transformation language, we decided to emulate some of the logical programming concepts from Prolog.
+We primarily included syntax concepts, ...?
+\textcolor{red}{don't wanna write anymore.
+write later} We also decided to include a timestamp at the end of each transformation to help with conflict resolution in a future stage where there may be multiple concurrent users making changes to the same set of data.
+Table \ref{table:transformations_summary} has a summary of how the structure of each data transformation will look like for each function.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | c | l | }
+ \hline
+    User Action & Data Transformation Syntax \\ \hline \hline
+    Add Row & addRow(tableName(columnName1:valuesUserPutIn1,\\ & columnName2:valuesUserPutIn2, ..., \\ & columnNameN:valuesUserPutInN),  timestamp)\\ & \\ \hline
+    Remove Row & removeRow(tableName(primaryKey1:primaryKeyValue1,\\ & primaryKey2:primaryKeyValue2, ..., \\ & primaryKeyN:primaryKeyValueN),  timestamp)\\ & \\ \hline
+    Add Column & addCol(tableName, nameOfNewCol, timestamp)\\ & \\ \hline
+    Remove Column & removeCol(tableName, nameOfColToRemove, timestamp)\\ & \\ \hline
+    Rename Column & renameCol(tableName, oldNameofCol, newNameOfCol, \\ & timestamp)\\ & \\ \hline
+    Edit Value & editValue(tableName(columnNameOfEditedValue1:\\ & editedValue1, columnNameOfEditedValue2:editedValue2, ..., \\ & columnNameOfEditedValueN:editedValueN),  timestamp)\\ 
+    \hline
+\end{tabular}
+\caption{The data transformations that will be logged to the transformation script as the user makes changes to the data.}
+\label{table:transformations_summary}
+\end{table}
+
+Sections \ref{sec:add_row}, \ref{sec:remove_row},  \ref{sec:add_col}, \ref{sec:remove_col}, \ref{sec:rename_col}, and \ref{sec:edit_value} will all use the same example scenario: a journalist intern has been tasked with writing a comparison article about the different animated movies produced in the last five years and how critic and user reviews differ.
+She has used her application to pull data from multiple movie information tables in a remote database.
+The resulting set of columns that she sees in her application will be referred to as a view.
+In this example, the columns being viewed are ``Movie Title'', and ``Critic Rating'', which come from the Movie Information table, and the ``User Rating'', ``Number of Likes'', and ``Website that user information was obtained from'' columns, which come from the User Information table.
+The Movie Information table has ``Movie Title'' as the primary key while the User Information table has ``Movie Title'' as its foreign key and ``Website that user information was obtained from'' as its primary key.
+Table \ref{table:add_row_table_example} shows what the initial data looks like after the user has pulled from the remote database.
+Note that schema mappings are also pulled to the application along with the data from the remote database.
+The schema mappings will be used to help determine what tables each of the respective columns in the view belong to.
+
+The following shows the table schema of the Movie Information and User Information tables:
+\begin{itemize}
+\item {Movie Information(Movie Title: String, Primary Key; Movie Debut Year: Int; Production Studio: String, NOT NULL; Total Gross: double; Critic Rating: String)}
+\item{User Information(Movie Title: String, Foreign Key; Website that user information was obtained from: String, Primary Key; User Rating: String; Number of Likes: double)}
+\end{itemize}
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | c | c | c | c | c | }
+ \hline
+    Movie Title & Critic  & User & Number & Website that user\\
+    & Rating & Rating & of Likes & information was \\
+    &&&& obtained from\\ \hline \hline
+    Despicable Me 2 & 5/5 & 5/5 & 220 000 & IMDB\\ \hline
+    Up & 5/5 & 5/5 & 225 000 & Facebook\\ \hline
+    Kung Fu Panda 2 & 4/5 & 5/5 & 500 000 & Rotten Tomatoes\\ \hline
+    Mars Needs Moms & 3/5 & 3/5 & 103 000 & Rotten Tomatoes\\
+    \hline
+\end{tabular}
+\caption{What the application displays to the user prior to the user making any changes to the data.}
+\label{table:add_row_table_example}
+\end{table}
+
+A constraint on the view is that the primary key(s) of each table have to be displayed in the view.
+The application must have that information as it potentially needs to have to be able to uniquely identify which value is being changed by the user (see Section \ref{sec:edit_value} for more details).
+Also, when a user decides to add a new row to the table (see Section \ref{sec:add_row} for more details), there needs to be values for the primary key(s) in order for a row to be added successfully.
+Allowing the user to choose which rows he/she wants does not guarantee that the primary key is chosen and this would mean that the user would lose certain functions without knowing why.
+Hence, when pulling information from each table, the application is required to pull all primary and foreign keys of the table.
+While we recognize that this will cause a lot of activity on the network and possibly take up much more space than required on the local machine, we could not think of a better solution.
+Due to this design decision, the example shown in Table \ref{table:add_row_table_example} does not include any details about whether the columns shown are the primary keys of their respective tables as we expect that the primary keys are already shown.
+There has been some discussion upon how to make it easier for the user to figure out which columns are primary keys and which columns come from which table but we did not feel that non-technical users could understand why certain columns were coloured the way they were.
+For users that can understand why columns were certain colours, it would certainly make it easier for them to perform actions such as add or remove a row (see Sections \ref{sec:add_row} and \ref{sec:remove_row}) but for the users who cannot figure out why the colours existed, it would only make it more distracting and confusing.
+Also, as we are working with the situation where there could possibly be third party apps accessing the database, we can only restrict what information gets pulled, we cannot really expect anything in terms of how the information is visualized.
+
+\subsubsection{Adding a row}
+\label{sec:add_row}
+When the user uses his/her application to add a row in the view, it is unclear about whether or not the user desires to include an additional row in all tables that have columns in the view.
+One of the design decisions we have decided on is to include a new row in the tables that have a value filled in at the column.
+
+In Table \ref{table:add_row_table_example2}, the user has just added in a new row and populated the ``Movie Title'' and ``Critic Rating'' columns with a value.
+If the user decides to not populate the row any further and push his/her changes, the remote database would have the Movie Information table add a new row with Megamind as the ``Movie Title'' value and 3/5 as the ``Critic Rating'' value.
+As the Movie Information table has other schema restrictions, namely the restriction that the ``Production Studio'' column has to have a value, there will be an error generated when the user chooses to push to the remote database.
+We have decided to let the database administrator for the remote database deal with issues like these.
+The alternative solution is to impose another restriction upon the application such that it will have to also pull columns that have schema specifications like UNIQUE or NOT NULL.
+However, we felt that it would be too confusing to the user to have such columns included.
+With primary/foreign keys, the information captured in those columns tend to be information that give the rest of the information in the row context and meaning.
+Columns that have schema specifications may not have such context and meaning and thus would be confusing to impose upon the user in their application.
+
+In Table \ref{table:add_row_table_example3}, the user has added more information about the movie ``Megamind''.
+As columns from the Movie Information table and the User Information table have been filled with values, there will be a new row added to the Movie Information and User Information table.
+Any issues with adding to the Movie Information table will have to be resolved by the local database administrator.
+If the user decides to try and add this row without a value for the ``Website that user has obtained rating from'' column, a primary key for the User Information table, the application will prompt the user to enter the value prior to allowing a database push.
+We did not feel that this was a strong or particularly confusing change to force the user to make and it would help reduce the number of issues pushed upon the database administrator.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | c | c | c | c | c | }
+ \hline
+    Movie Title & Critic  & User & Number & Website that user\\
+    & Rating & Rating & of Likes & information was \\
+    &&&& obtained from\\ \hline \hline
+    Despicable Me 2 & 5/5 & 5/5 & 220 000 & IMDB\\ \hline
+    Up & 5/5 & 5/5 & 225 000 & Facebook\\ \hline
+    Kung Fu Panda 2 & 4/5 & 5/5 & 500 000 & Rotten Tomatoes\\ \hline
+    Mars Needs Moms & 3/5 & 3/5 & 103 000 & Rotten Tomatoes\\ \hline
+    Megamind & 3/5 &  &  & \\
+    \hline
+\end{tabular}
+\caption{The user is trying to add a new row to the table where only columns from the Movie Information table is filled.}
+\label{table:add_row_table_example2}
+\end{table}
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{ | c | c | c | c | c | }
+ \hline
+    Movie Title & Critic  & User & Number & Website that user\\
+    & Rating & Rating & of Likes & information was \\
+    &&&& obtained from\\ \hline \hline
+    Despicable Me 2 & 5/5 & 5/5 & 220 000 & IMDB\\ \hline
+    Up & 5/5 & 5/5 & 225 000 & Facebook\\ \hline
+    Kung Fu Panda 2 & 4/5 & 5/5 & 500 000 & Rotten Tomatoes\\ \hline
+    Mars Needs Moms & 3/5 & 3/5 & 103 000 & Rotten Tomatoes\\ \hline
+    Megamind & 3/5 & & 200 000 & IMDB \\
+    \hline
+\end{tabular}
+\caption{The user is trying to fill in a new row that she has added. Columns from the Movie Information and User Information table have values added to them.}
+\label{table:add_row_table_example3}
+\end{table}
+
+When the user adds a row, the change that is logged to the transformation script has the following structure: addRow(tableName(columnName1:valuesUserPutIn1, columnName2:valuesUserPutIn2, ..., columnNameN:valuesUserPutInN), timestamp).
+
+\subsubsection{Removing a row}
+\label{sec:remove_row}
+When removing a row, the user has to delete all values from columns that belong to a single table before the row can be successfully deleted in the database.
+This design decision was made to avoid users accidentally deleting whole rows when they just want to delete a single value from a column.
+We thought that this reflects how a user would normally use the system \textemdash if he/she really wanted to delete a row, they would go delete the whole row already.
+As we felt that this accurately reflected a normal usage situation, we did not consider this as a constraint that imposed extra annoyance upon the user.
+
+When the user removes a row, the change that is logged to the transformation script has the following structure:
+removeRow(tableName(primaryKey1:\\primaryKeyValue1, primaryKey2:primaryKeyValue2, ..., primaryKeyN:\\primaryKeyValueN), timestamp).
+We are just using the primary key to identify which specific row to delete and have decided to let the database system handle delete propagation (i.e., deleting the rows that other tables have which use values referencing the ones in the deleted row).
+
+\subsubsection{Adding a column}
+\label{sec:add_col}
+When adding a column, there was a choice between having a column added to all tables that have columns in the current view or having the column added to one of the tables that had columns in the view.
+Originally we had the column added to all tables that have columns in the current view but quickly realized that it would lead to a lot of update problems along with the fact that it would take up extra memory.
+In the end, we decided to have the column added to the table with the most columns in the current view.
+If tables have an equal number of columns in the view, the table can be chosen in one of three ways:
+\begin{itemize}
+\item {Arbitrarily/at random}
+\item{Based on the name of the table (e.g., add the column to the table with the name that is listed first when table names are sorted alphanumerically)}
+\item{Based on the statistics of table usage that the database stores (e.g., the column could be added to the table that is queried more often).
+However, this would involve some back and forth dialog between the database and the remote application which may not be a good idea for performance or database security measures}
+\end{itemize}
+
+We decided to choose the second method to resolve any possible conflicts between table choice when adding columns.
+The add column change is logged as follows: addCol(tableName, nameOfNewCol, timestamp).
+
+\subsubsection{Removing a column}
+\label{sec:remove_col}
+While discussing this function, there was some concern about whether or not users should be allowed to remove a column and if they were, whether they should be allowed to remove primary key columns.
+In the end, we decided to only allow the removal of non-primary key columns for database security reasons.
+It could be argued that the user should be able to do whatever he/she wishes upon the data as that is the point of enabling users from all backgrounds to use and update data but upon weighing the pros and cons of this issue, it was decided that it was too easy for the user to cause major damage to the data source if primary key removal was allowed.
+If the user wishes to remove a primary key column, he/she could contact the database administrator.
+The remove column change would be logged into the transformation script as follows: removeCol(tableName, nameOfColToRemove, timestamp).
+
+\subsubsection{Renaming a column}
+\label{sec:rename_col}
+Although the situation where a user would rename a column is uncommon, we decided to enable support for this use case as it could potentially come up.
+The transformation script syntax for this change is: renameCol(tableName, oldNameofCol, newNameOfCol, timestamp) where the tableName is the name of the table that the column to be renamed is located in, oldNameOfCol is the old name of the column used to identify which column to rename, and newNameOfCol is the new name of the column.
+
+\subsubsection{Editing a value}
+\label{sec:edit_value}
+When editing a value, a user can make either a 1-1 or 1-0 change (i.e., a 1-1 change is when the user can change one value to another while a 1-0 change would be when a user deletes the value).
+As with the add row functionality (see Section \ref{sec:add_row}), if a user ends up violating schema constraints when editing a value, the database administrator will have to approve/deal with the changes.
+Some of the possible schema violations include changing a unique value to a non-unique one when the column specifies that it needs to have unique values or deleting a value from a column that has specified that it cannot be null.
+The edit a value change would be logged into the transformation script as follows: editValue(tableName(columnNameOfEditedValue1: editedValue1, columnNameOfEditedValue2:editedValue2, ..., columnNameOfEditedValueN:editedValueN).
 
 \subsection{View Maintenance}
-View maintenance \textcolor{red}{cite stuff here}
+View maintenance is also something that factored into our considerations throughout the project.
+There were long discussions about whether or not we should allow the user to choose when updates would be pushed to the remote database or if the application should choose.
+In the end, we decided on a mixed-initiative approach where the user would push if certain types or a certain number of changes have occurred while also allowing the user to choose when to push changes.
+The primary concern was that schema changes (see Section \ref{sec:add_row}, \ref{sec:remove_row}, \ref{sec:add_col}, and \ref{sec:remove_col}) were important enough to push right away as the longer the wait period, the potential for concurrent users becomes greater.
+Value changes are considered less important; although value changes from one user may conflict with another user, it was decided to use the timestamp to resolve conflicts.
+The application can push to the remote data source after a certain number of value changes.
 
@@
 \subsection{Transformation Languages}
-There is no lack of data transformation tools
+While there are an abundance of data transformation tools \cite{raman2001potter, kandel2011wrangler}, these tools often focus on visualizing the data transformation rather than the transformation language used to represent the data changes behind the scenes.
+
+- something about how logical programming concepts popular with being used to declaratively represent data transformations
+- find papers about this
+- stuff
 
@@
 
+This problem can also be investigated from an information visualization/HCI perspective.
+The best way to present information to the user to help facilitate use and understanding of how the data can be manipulated could be examined.
+However, as stated before, this is more on the application side of this problem which we are not specifically dealing with in the scope of this project.
+
 \bibliographystyle{abbrv}

### 2015-04-14 Final_Paper/Paper.tex
What we have now.

@@
 \usepackage{color}
-\usepackage{tabularx}
+\usepackage[margin=1.0in]{geometry}
 
@@
 \subsection{Types of Changes Supported}
+\label{sec:types_of_changes}
 Based on the various classes of changes mentioned in Wrangler \cite{kandel2011wrangler}, we decided to focus on two of them: map and reshape.
@@
 \item {Arbitrarily/at random}
-\item{Based on the name of the table (e.g., add the column to the table with the name that is listed first when table names are sorted alphanumerically)}
-\item{Based on the statistics of table usage that the database stores (e.g., the column could be added to the table that is queried more often).
+\item {Based on the name of the table (e.g., add the column to the table with the name that is listed first when table names are sorted alphanumerically)}
+\item {Based on the statistics of table usage that the database stores (e.g., the column could be added to the table that is queried more often).
 However, this would involve some back and forth dialog between the database and the remote application which may not be a good idea for performance or database security measures}
@@
 The application can push to the remote data source after a certain number of value changes.
+As for quantifying the number of changes that should occur before the application automatically pushes to the data source, that has yet to be determined.
 
 \subsection{Proof of Concept}
-After designing the transformation language, we decided to implement the scenario workflow in order to show that our transformation language and design decisions could work to update a data source.
+After designing the transformation language, we decided to implement the scenario in order to show that our transformation language and design decisions could work to update a data source.
 The only extra step that was not included in \cite{arniThesis} was to allow for the user to choose which changes to sync to the remote data source.
-\textcolor{red}{add screen shots!!
+We created an interface using Netbeans and \textcolor{red}{add screen shots!!
 Kailang can write about all the cool things she did here!!
@@
 \section{Related Work}
-\textcolor{red}{need stuff here can probably reuse the stuff from the project update but will need more + editing}
+\textcolor{red}{need stuff here can probably reuse the stuff from the project update but will need more + editing} \\
 \textbf{\textcolor{green}{Stuff in this section is ripped from the project update}}
-One of the largest problems that we foresee is creating a transformation language to track all the possible data changes as well as how to propagate those changes correctly from one source to another.
-Hence, most of our related work focuses on systems that are similar or areas of work where we feel we can draw inspiration from how they tackled a similar problem to the one we are facing.
 
-\subsection{Data Sharing Systems}
-Data sharing situations are not new; there have been many collaborative data sharing systems (CDSS) \cite{ives2008orchestra, ives2005orchestra, karvounarakis2013collaborative}and data transformation language \cite{kandel2011wrangler, lakshmanan2001schemasql} created to try and address this issue.
-From CDSS systems, we believe we can examine how it handles schema mapping change propagations to inform our own design as this is exactly the same problem we are facing.
+\subsection{ETL: Extract, Transform, Load}
 
 \subsection{Transformation Languages}
-While there are an abundance of data transformation tools \cite{raman2001potter, kandel2011wrangler}, these tools often focus on visualizing the data transformation rather than the transformation language used to represent the data changes behind the scenes.
+While there has been a large body of work done on various data transformation tools \cite{kandel2011wrangler, raman2001potter, kandel2012profiler}, these tools often focus on visualizing the data transformation rather than the transformation language used to represent the data changes behind the scenes.
+In fact, most of the work in the data transformation language domain seems to be classified into one of two categories.
+Either the work focuses on trying to help uses visualize the data transformations that have occurred \cite{kandel2011wrangler, raman2001potter, kandel2012profiler} or it focuses on modifying existing transformation languages to help with the data transformation process \cite{ilprints409, lakshmanan2001schemasql}.
+Due to the lack of literature about the specifics of data transformation languages, we decided to cast our net further and examine data transformation language frameworks.
+There was a minimal amount of literature on using logical programming concepts to create a data transformation language \textcolor{red}{need citation}.
+The idea was that people find declarative languages like logical programming languages easier to understand as it focuses on the relationships between the data models rather than the method of how to locate a certain piece of information to work with \textcolor{red}{need citation}.
+There has been research working with declarative data transformation languages for different data models \cite{bry2002towards, lawley2006practical, tarau2009embedded} so we decided to try and model our data transformation language accordingly.
 
-- something about how logical programming concepts popular with being used to declaratively represent data transformations
-- find papers about this
-- stuff
+\textcolor{red}{incorporate into paragraph}
+\cite{bry2002towards} : declarative data transformation language for XML and semi-structured data
+
+\cite{lawley2006practical} : implementation of a pattern based language for transforming data models
+
+\cite{tarau2009embedded} : embedded logical language framework
+
+\subsection{Supported Changes}
+Another part of the system that we had to figure out was the actual scenarios that would be executed on our system.
+We decided to do a literature search to try and figure out what our system had to support \cite{galhardas:inria-00072476, kandel2011wrangler, kandel2012profiler,rahm2000data, raman2001potter}.
+From this requirements solicitation, we determined that our data transformation language had to focus on the logical transformations rather than the physical (this is handled by the fact that our data transformation language borrows from logical programming concepts) and that the types of changes we had to support were map and reshape changes (see Section \ref{sec:types_of_changes}) \cite{kandel2011wrangler}.
 
@@
 Currently we have chosen to only allow data from a single table be viewed in our interface.
-However, given that this project could be extended in the future to include viewing data that is from multiple tables, we are looking ahead at literature in the view maintenance area to try and predict what challenges our interface might face \cite{ agrawal1997efficient, agrawal2009asynchronous, gupta1999materialized, zhou2007lazy}.
+However, given that this project could be extended in the future to include viewing data that is from multiple tables, we are looking ahead at literature in the view maintenance domain to try and predict what challenges our interface might face \cite{ agrawal1997efficient, agrawal2009asynchronous, gupta1999materialized, zhou2007lazy}.
 We also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.

### 2015-04-14 Final_Paper/Paper.tex
Paper.

@@
 \usepackage[margin=1.0in]{geometry}
+\newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
 
@@
 
+\noindent \textcolor{blue}{do we want to put the examples here?}
+
 \noindent Our primary contributions are:
@@
 	\item{Defined a data transformation language using logical programming concepts to capture changes made to a relational data model}
-	\item{Design decisions about how the relational model should react and handle schema changes}
-	\item{Preliminary investigation into view maintenance policies for the remote datastore}
-	\item{Further detailing the scenario in which an application views a subset of columns from one or more tables in a remote dataset}
+	\item{Proposed how the relational model could react and handle schema changes}
+	\item{Investigated view maintenance policies for the remote datastore}
+	\item{Detailed scenarios in which an application views a subset of columns from one or more tables in a remote dataset}
 	\item{Decided which issues to assign to the server side and which to the client side}
@@
 \section{Motivation}
-In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables, it is important to figure out how this relates to the real world.
-The example in \cite{arniThesis} talks about a sports reporter who retrieves select information about basketball players to do an article on.
-In this case, the reporter only takes out columns regarding a player's name, his high school, and the first team he played on after entering the NBA.
-Compared to all the information that is stored about NBA players on a main database (e.g., salary, weight, height, number of transfers, agent name, etc.), the reporter has only taken a very small subset of columns.
-Another example could be a school administrator who wants to decide which students to give awards to purely based on academic merit.
-He/she could go to the main database storing all student information (e.g., the database could have information regarding a student's name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) and only choose to view the student number and GPA column so that he/she will not be biased towards any names or gender.
+In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables.
+It is important to figure out how this relates to the real world.
+For instance, a sports reporter wishes to write an article on a select number of basketball players and retrieves a specific set of information related to the chosen players \cite{arniThesis}.
+In this case, the reporter only uses columns such as player's name, his high school, and the first team he played for after entering the NBA.
+Compared to all the information that is stored about NBA players on a main database (e.g., salary, weight, height, number of transfers, agent name, etc.), the reporter has only looked a small subset of columns.
+
+Consider another example: at the end of a school year, a school administrator is deciding the students to whom academic merit prizes are to be awarded.
+He/she could go to the main database storing all student information (e.g., the database stores detailed student information including a name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) but only wishes to access student number and GPA columns to ensure that he/she will not be biased towards any name or gender.
 
@@
 In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application.
-This was a movie star dataset that had the columns ``Star Id'', ``Name'', and ``Gender''.
-After supplementing some additional data to the dataset, namely the columns ``Movie Title'', ``Production Studio'', and ``Critic Rating'', we used this dataset.
-We also added in a second table detailing user information with the columns ``Movie Title'' which is a foreign key referencing the ``Movie Title'' column in the Movie Information table, ``Website that user information was obtained from'', ``User Rating'', and ``Number of Likes''.
+\hilight{This was a movie star dataset that had the columns ``Star Id'', ``Name'', and ``Gender''.
+After supplementing some additional data to the dataset, namely the columns ``Movie Title'', ``Production Studio'', and ``Critic Rating'', we used this dataset.} \textcolor{blue} {awkward wording} We also added in a second table detailing user information with the columns ``Movie Title'' which is a foreign key referencing the ``Movie Title'' column in the Movie Information table, ``Website that user information was obtained from'', ``User Rating'', and ``Number of Likes''.
 
 \section{Solution}
-The basic workflow of the scenario described in Section \ref{sec:introduction} will remain the same.
+The basic workflow of the scenario described in Section \ref{sec:introduction} remains the same.
 However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformations increase along with the amount of information that must be tracked (see Section \ref{sec:transformation_language}).
@@
 \section{Related Work}
-\textcolor{red}{need stuff here can probably reuse the stuff from the project update but will need more + editing} \\
-\textbf{\textcolor{green}{Stuff in this section is ripped from the project update}}
-
-\subsection{ETL: Extract, Transform, Load}
-
 \subsection{Transformation Languages}
@@
 Due to the lack of literature about the specifics of data transformation languages, we decided to cast our net further and examine data transformation language frameworks.
-There was a minimal amount of literature on using logical programming concepts to create a data transformation language \textcolor{red}{need citation}.
-The idea was that people find declarative languages like logical programming languages easier to understand as it focuses on the relationships between the data models rather than the method of how to locate a certain piece of information to work with \textcolor{red}{need citation}.
-There has been research working with declarative data transformation languages for different data models \cite{bry2002towards, lawley2006practical, tarau2009embedded} so we decided to try and model our data transformation language accordingly.
-
-\textcolor{red}{incorporate into paragraph}
-\cite{bry2002towards} : declarative data transformation language for XML and semi-structured data
-
-\cite{lawley2006practical} : implementation of a pattern based language for transforming data models
-
-\cite{tarau2009embedded} : embedded logical language framework
+There was a minimal amount of literature on using logical programming concepts to create a data transformation language \cite{bry2002towards}.
+The idea was that people find declarative languages like logical programming languages easier to understand as it focuses on the relationships between the data models rather than the method of how to locate a certain piece of information to work with \cite{bry2002towards, tarau2009embedded}.
+There has been research working with declarative data transformation languages for different data models, highlighting the flexibility requirements in order to keep current with today's less structured data, \cite{bry2002towards, lawley2006practical, tarau2009embedded} so we decided to try and model our data transformation language accordingly.
 
@@
 \subsection{View Maintenance}
-Currently we have chosen to only allow data from a single table be viewed in our interface.
-However, given that this project could be extended in the future to include viewing data that is from multiple tables, we are looking ahead at literature in the view maintenance domain to try and predict what challenges our interface might face \cite{ agrawal1997efficient, agrawal2009asynchronous, gupta1999materialized, zhou2007lazy}.
-We also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.
+As our application takes a mixed-initiative approach to updating the remote data source, we also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.
+Through looking at how views and indices are maintained in a large distributed database \cite{agrawal1997efficient}, we realized that many of the design decisions we made for our transformation language (see Section \ref{sec:transformation_language}) fit quite well with the model suggested in the paper.
+For example, their discussion on the different ways to push updates and the decisions made on whether the system or the application should handle certain update issues do mirror our decisions.
 
@@
 
+The update policy of this system can be examined more closely to see if we can draw upon more of the principles and ideas discussed in the distributed database-application system PNUTS \cite{agrawal1997efficient}.
+
 \bibliographystyle{abbrv}

### 2015-04-14 Final_Paper/Paper.tex
Complete version

@@
 \usepackage{color}
+\usepackage{graphicx}
 \usepackage[margin=1.0in]{geometry}
 \newcommand{\hilight}[1]{\colorbox{yellow}{#1}}
+\newcommand{\figref}[1]{Figure \ref{#1}}
 
@@
 \label{sec:introduction}
-This project is built on top of the theoretical framework done in the M.Sc thesis of Arni Thrastarson \cite{arniThesis} and focuses on the first of the four scenarios detailed in the work.
+This project builds on the theoretical framework described in the M.Sc thesis of Arni Thrastarson \cite{arniThesis} and focuses on the first of the four scenarios detailed in the work.
 The first scenario investigates how changes made by a data application can be captured and propagated to a remote data source.
-The main workflow likely involves a (possibly third party) application pulling the data from a remote data source to use locally.
-The application creates a local copy of the data along with a schema mapping to identify the source table(s) of the data attributes.
-A user using the application may notice inconsistencies with the data and correct them.
-Any changes made are logged in a transformation script and when the time comes to update the data source, the transformation script is sent to the data store where the changes detailed in the script are applied.
+The main workflow of this scenario involves a (possibly third party) application pulling the data from a remote data source to use locally.
+In order to facilitate updates to the data source, the application receives a schema mapping to identify the source table(s) of the data attributes along with a copy of the data it is requesting.
+The user using the application after retrieving the data may choose to add, remove, or edit the data.
+These changes are then logged to a transformation script which is later used to update the main copy of the data in the remote data store.
 
-In this project, we focus on three different ways a user can use the application to fetch data. For the rest of this paper, we shall call each way of data fetching a sub-scenario. 
+In this project, we focus on three different ways a user can use the application to fetch data. For the rest of this paper, we shall call each of the following a sub-scenario. 
 \begin{enumerate}
@@
 	\item {When the user chooses to view a subset of the columns in one table}
-	\item {When the user chooses to view a subset of the columns in multiple tables}
+	\item {When the user chooses to view a subset of columns in multiple tables}
 \end{enumerate}
-
-\noindent \textcolor{blue}{do we want to put the examples here?}
 
@@
 \begin{enumerate}
-	\item{Defined a data transformation language using logical programming concepts to capture changes made to a relational data model}
-	\item{Proposed how the relational model could react and handle schema changes}
-	\item{Investigated view maintenance policies for the remote datastore}
-	\item{Detailed scenarios in which an application views a subset of columns from one or more tables in a remote dataset}
-	\item{Decided which issues to assign to the server side and which to the client side}
+	\item{Defining a data transformation language using logical programming concepts to capture changes made to a relational data model}
+	\item{Proposing how the data source could react and handle schema changes}
+	\item{Investigating view maintenance policies for the remote data source}
+	\item{Detailing scenarios in which an application views a subset of columns from one or more tables in a remote data source}
+	\item{Deciding on the division of responsibility between the database administrator and the application user in regards to data updates}
 \end{enumerate}
@@
 \section{Motivation}
-In our project, our main focus is on a scenario where the user views any subset of columns from one or more tables.
-It is important to figure out how this relates to the real world.
-For instance, a sports reporter wishes to write an article on a select number of basketball players and retrieves a specific set of information related to the chosen players \cite{arniThesis}.
-In this case, the reporter only uses columns such as player's name, his high school, and the first team he played for after entering the NBA.
-Compared to all the information that is stored about NBA players on a main database (e.g., salary, weight, height, number of transfers, agent name, etc.), the reporter has only looked a small subset of columns.
+\label{sec:motivation}
+We believe that the three sub-scenarios outlined are realistic.
+The first sub-scenario can be easily envisioned when thinking about a librarian trying to re-catalogue all the books in a library.
+The librarian would need to have all the information about a book in order to catalogue and double check the information.
 
-Consider another example: at the end of a school year, a school administrator is deciding the students to whom academic merit prizes are to be awarded.
-He/she could go to the main database storing all student information (e.g., the database stores detailed student information including a name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) but only wishes to access student number and GPA columns to ensure that he/she will not be biased towards any name or gender.
+As a motivating example for using a subset of one table, we consider a school administrator who is deciding to which student to award an academic merit prize.
+The school administrator could go to the main database storing all student information (e.g., name, student number, GPA, guardian contact information, date of birth, allergy/medication information, family doctor contact, etc.) but only wishes to access student number and GPA columns to ensure that the prize is awarded fairly.
 
-An example where a user might need to view a subset of columns in multiple tables would be if the user is trying to find detailed information about proteins found in different microbiological organisms.
-The microbiological organisms and the various proteins it produces could be stored in one table while detailed information about each protein is stored in another.
-If a user wanted to see certain characteristics of a protein found in specific organisms, he/she would have use a subset of columns from both the microbiological organism table along with the protein time to get all the relevant information.
+Lastly, users might need to view a subset of columns in multiple tables if they are a journalist intern who has been tasked with compiling a report for an online journal.
+The report could be on differences in critic ratings and user ratings for animated movies produced in the last five years which would require the intern to look at information from a movies and user information table..
 
@@
 In our project, we decided to borrow the dataset used in CPSC 304, an introductory relational database course at UBC, to use as the test dataset in our application.
-\hilight{This was a movie star dataset that had the columns ``Star Id'', ``Name'', and ``Gender''.
-After supplementing some additional data to the dataset, namely the columns ``Movie Title'', ``Production Studio'', and ``Critic Rating'', we used this dataset.} \textcolor{blue} {awkward wording} We also added in a second table detailing user information with the columns ``Movie Title'' which is a foreign key referencing the ``Movie Title'' column in the Movie Information table, ``Website that user information was obtained from'', ``User Rating'', and ``Number of Likes''.
+Specifically we use the ``MOVIESTAR'' table, which has three columns ``STARID'', ``NAME'', and ``GENDER''.
+After supplementing some additional data to the dataset, namely the columns ``MOVIE TITLE'', ``PRODUCTION STUDIO'', and ``CRITIC RATING'', we used this dataset.
+We then created a second table detailing user information using the column ``Movie Title'' as a foreign key to reference the ``MOVIE TITLE'' column in the Movie Information table, ``SOURCE WEBSITE OF USER INFORMATION'', ``USER RATING'', and ``NUMBER OF LIKES''.
 
@@
 The basic workflow of the scenario described in Section \ref{sec:introduction} remains the same.
-However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformations increase along with the amount of information that must be tracked (see Section \ref{sec:transformation_language}).
-In this project, we have decided to support the following user changes: adding a column, removing a column, editing a column name, changing a table cell's current value, removing a table cell's value to make it null, adding a row, and removing a row.
+However, as the user obtains more flexibility over what columns can be chosen to view in his/her application,  the complexity of the data transformation increases due to the amount of information that must be tracked (see Section \ref{sec:transformation_language}).
+In this project, we have decided to support the following user changes: adding a column, removing a column, renaming a column, editing a table cell's value, removing a table cell's value, adding a row, and removing a row.
 Any changes that cause a violation in schema definitions (e.g., removing a row's primary key) will be resolved on the server side by the database administrator.
-This design decision was due to not wanting to restrict the types of users that would be able to use an app to find data to examine.
+This design decision was due to not wanting to restrict the types of users that would be able to use an application to explore and update data.
 As we have to assume that there could be users unfamiliar with schema/storage restrictions, we decided to ensure that schema violations are resolved by someone with the technical understanding of what error has occurred.
@@
 \label{sec:types_of_changes}
-Based on the various classes of changes mentioned in Wrangler \cite{kandel2011wrangler}, we decided to focus on two of them: map and reshape.
-Map changes deal with "[transforming] one input data row to zero, one, or multiple output rows" \cite{kandel2011wrangler}.
+We have decided to focus on two of the eight changes mentioned in Wrangler \cite{kandel2011wrangler}: map and reshape.
+Map changes deal with ``[transforming] one input data row to zero, one, or multiple output rows'' \cite{kandel2011wrangler}.
 This essentially means that the row changes that are supported in our application (e.g., add row, delete row, edit value(s) in row(s)) all fall within this class of changes.
@@
 \label{sec:transformation_language}
-As we designed the transformation language, we realized it did not make sense to work off the simplest of our three sub-scenarios as the transformation language did not adapt well when the complexity of the sub-scenario was increased.
-Hence, we decided to work off the most complicated of our three sub-scenarios, the situation where an application has pulled a subset of columns from more than one table, and design our transformation language from that point onwards.
-We then retroactively checked to ensure that the transformations worked for the simpler sub-scenarios.
+During the design of the transformation language, we realized that it did not make sense to start with the simplest of our three sub-scenarios as the transformation language did not scale well with increasing complexity.
+Therefore, we are basing our decisions around the most complicated of our three sub-scenarios: the situation where an application has pulled a subset of columns from more than one table.
+To ensure that the transformations worked for the simpler sub-scenarios, every step was aligned with all consequences in mind.
 
-When designing the data transformation language, we decided to emulate some of the logical programming concepts from Prolog.
-We primarily included syntax concepts, ...?
-\textcolor{red}{don't wanna write anymore.
-write later} We also decided to include a timestamp at the end of each transformation to help with conflict resolution in a future stage where there may be multiple concurrent users making changes to the same set of data.
+When designing the data transformation language, we decided to emulate some of the logical programming concepts from Prolog (none too biased by our instructor's well-known preference for Datalog).
+We primarily borrowed from the syntactical style and the easy-to-understand format.
+The addition of a timestamp was in preparation for a possible future extension of the project where multiple concurrent users would make updates to the data source.
 Table \ref{table:transformations_summary} has a summary of how the structure of each data transformation will look like for each function.
 
-\begin{table}[h!]
+\begin{table}[!ht]
 \centering
@@
     User Action & Data Transformation Syntax \\ \hline \hline
-    Add Row & addRow(tableName(columnName1:valuesUserPutIn1,\\ & columnName2:valuesUserPutIn2, ..., \\ & columnNameN:valuesUserPutInN),  timestamp)\\ & \\ \hline
-    Remove Row & removeRow(tableName(primaryKey1:primaryKeyValue1,\\ & primaryKey2:primaryKeyValue2, ..., \\ & primaryKeyN:primaryKeyValueN),  timestamp)\\ & \\ \hline
+    Add Row & addRow(tableName(columnName1:valuesUserPutIn1,\\ & columnName2:valuesUserPutIn2, \\ & ..., \\ & columnNameN:valuesUserPutInN),  timestamp)\\ & \\ \hline
+    Remove Row & removeRow(tableName(primaryKey1:primaryKeyValue1,\\ & primaryKey2:primaryKeyValue2, \\ & ..., \\ & primaryKeyN:primaryKeyValueN),  timestamp)\\ & \\ \hline
     Add Column & addCol(tableName, nameOfNewCol, timestamp)\\ & \\ \hline
@@
     Rename Column & renameCol(tableName, oldNameofCol, newNameOfCol, \\ & timestamp)\\ & \\ \hline
-    Edit Value & editValue(tableName(columnNameOfEditedValue1:\\ & editedValue1, columnNameOfEditedValue2:editedValue2, ..., \\ & columnNameOfEditedValueN:editedValueN),  timestamp)\\ 
+    Edit Value & editValue(tableName(originalPrimaryKey, ..., originalPrimaryKeyN,\\ & columnNameOfEditedValue1:editedValue1, \\ & columnNameOfEditedValue2:editedValue2, \\ & ..., columnNameOfEditedValueN:editedValueN),  timestamp)\\
     \hline
 \end{tabular}
-\caption{The data transformations that will be logged to the transformation script as the user makes changes to the data.}
+\caption{The data transformations logged to the transformation script as the user makes changes to the data.}
 \label{table:transformations_summary}
@@
 
-Sections \ref{sec:add_row}, \ref{sec:remove_row},  \ref{sec:add_col}, \ref{sec:remove_col}, \ref{sec:rename_col}, and \ref{sec:edit_value} will all use the same example scenario: a journalist intern has been tasked with writing a comparison article about the different animated movies produced in the last five years and how critic and user reviews differ.
-She has used her application to pull data from multiple movie information tables in a remote database.
-The resulting set of columns that she sees in her application will be referred to as a view.
-In this example, the columns being viewed are ``Movie Title'', and ``Critic Rating'', which come from the Movie Information table, and the ``User Rating'', ``Number of Likes'', and ``Website that user information was obtained from'' columns, which come from the User Information table.
-The Movie Information table has ``Movie Title'' as the primary key while the User Information table has ``Movie Title'' as its foreign key and ``Website that user information was obtained from'' as its primary key.
-Table \ref{table:add_row_table_example} shows what the initial data looks like after the user has pulled from the remote database.
+\noindent The rest of this paper will use the following tables:
+\begin{itemize}
+\item {Movie Information(Movie Title: String, Primary Key; Movie Debut Year: Int; Production Studio: String, not null; Total Gross: double; Critic Rating: String)}
+\item{User Information(Movie Title: String, Foreign Key; Website that user information was obtained from: String, Primary Key; User Rating: String; Number of Likes: double)}
+\end{itemize}
+
+Once an application has pulled data from a remote database, the resulting set of columns shown in the application will be referred to as a view.
+For the journalist intern example mentioned in Section \ref{sec:motivation}, the columns being viewed are ``MOVIE TITLE'', and ``CRITIC RATING'', both of which come from the Movie Information table, ``USER RATING'', ``NUMBER OF LIKES'', and ``SOURCE WEBSITE OF USER INFORMATION'' columns, which come from the User Information table.
+The Movie Information table has ``MOVIE TITLE'' as the primary key while the User Information table has ``MOVIE TITLE'' as its foreign key and ``SOURCE WEBSITE OF USER INFORMATION'' as its primary key.
+Table \ref{table:add_row_table_example} shows what the displayed view after the user has pulled from the remote database.
 Note that schema mappings are also pulled to the application along with the data from the remote database.
@@
 
-The following shows the table schema of the Movie Information and User Information tables:
-\begin{itemize}
-\item {Movie Information(Movie Title: String, Primary Key; Movie Debut Year: Int; Production Studio: String, NOT NULL; Total Gross: double; Critic Rating: String)}
-\item{User Information(Movie Title: String, Foreign Key; Website that user information was obtained from: String, Primary Key; User Rating: String; Number of Likes: double)}
-\end{itemize}
-
-\begin{table}[h!]
+\begin{table}[!ht]
 \centering
@@
 \end{tabular}
-\caption{What the application displays to the user prior to the user making any changes to the data.}
+\caption{An example user-application display prior to the user making any changes to the data.}
 \label{table:add_row_table_example}
@@
 
-A constraint on the view is that the primary key(s) of each table have to be displayed in the view.
-The application must have that information as it potentially needs to have to be able to uniquely identify which value is being changed by the user (see Section \ref{sec:edit_value} for more details).
+A constraint on the view is that the primary key(s) of each table must be displayed.
+This is a necessary detail as the primary key is needed to be able to uniquely identify the value being changed by the user (see Section \ref{sec:edit_value} for more details).
 Also, when a user decides to add a new row to the table (see Section \ref{sec:add_row} for more details), there needs to be values for the primary key(s) in order for a row to be added successfully.
-Allowing the user to choose which rows he/she wants does not guarantee that the primary key is chosen and this would mean that the user would lose certain functions without knowing why.
-Hence, when pulling information from each table, the application is required to pull all primary and foreign keys of the table.
-While we recognize that this will cause a lot of activity on the network and possibly take up much more space than required on the local machine, we could not think of a better solution.
+Allowing the user to choose which rows he/she wants does not guarantee that the primary key is chosen and this may result in a seemingly inexplicable loss of certain application functions.
+Therefore, when pulling information from each table, the application is required to pull all primary and foreign keys of the table.
+While we recognize that this will cause a lot of activity on the network and possibly take up more local memory, we could not think of a better solution.
 Due to this design decision, the example shown in Table \ref{table:add_row_table_example} does not include any details about whether the columns shown are the primary keys of their respective tables as we expect that the primary keys are already shown.
-There has been some discussion upon how to make it easier for the user to figure out which columns are primary keys and which columns come from which table but we did not feel that non-technical users could understand why certain columns were coloured the way they were.
-For users that can understand why columns were certain colours, it would certainly make it easier for them to perform actions such as add or remove a row (see Sections \ref{sec:add_row} and \ref{sec:remove_row}) but for the users who cannot figure out why the colours existed, it would only make it more distracting and confusing.
-Also, as we are working with the situation where there could possibly be third party apps accessing the database, we can only restrict what information gets pulled, we cannot really expect anything in terms of how the information is visualized.
+There has been some discussion regarding how to make it easier for the user to determine primary key columns as opposed to other columns but we did not feel that non-technical users could understand the difference intuitively.
+For users that can understand why the idea of different columns belonging to different tables, it would certainly make it easier for them to perform actions such as add or remove a row (see Sections \ref{sec:add_row} and \ref{sec:remove_row}) but for the very non-technical users, it would only make it more distracting and confusing.
+Also, as we are possibly working with third party applications, we can only restrict what information gets pulled, we cannot expect anything in terms of how the information is visualized to the users.
 
@@
 \label{sec:add_row}
-When the user uses his/her application to add a row in the view, it is unclear about whether or not the user desires to include an additional row in all tables that have columns in the view.
-One of the design decisions we have decided on is to include a new row in the tables that have a value filled in at the column.
+When the user uses his/her application to add a row in the view, it is unclear whether or not the user desires to include an additional row in all tables that have columns in the view.
+One of the design factors we have decided on is to include a new row in the tables that have a value filled in at the column.
 
-In Table \ref{table:add_row_table_example2}, the user has just added in a new row and populated the ``Movie Title'' and ``Critic Rating'' columns with a value.
-If the user decides to not populate the row any further and push his/her changes, the remote database would have the Movie Information table add a new row with Megamind as the ``Movie Title'' value and 3/5 as the ``Critic Rating'' value.
-As the Movie Information table has other schema restrictions, namely the restriction that the ``Production Studio'' column has to have a value, there will be an error generated when the user chooses to push to the remote database.
-We have decided to let the database administrator for the remote database deal with issues like these.
-The alternative solution is to impose another restriction upon the application such that it will have to also pull columns that have schema specifications like UNIQUE or NOT NULL.
-However, we felt that it would be too confusing to the user to have such columns included.
-With primary/foreign keys, the information captured in those columns tend to be information that give the rest of the information in the row context and meaning.
-Columns that have schema specifications may not have such context and meaning and thus would be confusing to impose upon the user in their application.
+In Table \ref{table:add_row_table_example2}, the user has just added in a new row and populated the ``MOVIE TITLE'' and ``CRITIC RATING'' columns with a value.
+If the user decides to not populate the row any further and push his/her changes, the remote database would have the Movie Information table add a new row with Megamind as the ``MOVIE TITLE'' value and 3/5 as the ``CRITIC RATING'' value.
+However, as the Movie Information table has other schema restrictions, namely that the ``PRODUCTION STUDIO'' column must be populated, there will be an error generated when the user chooses to push to the remote database.
+We have decided to let the database administrator deal with issues like these at the remote database end.
+The alternative solution is to impose another restriction upon the application such that it will have to also pull columns that have schema specifications like ``unique'' or ``not null''.
+However, we felt that it would be too confusing for the end-user to have such columns included.
+With primary/foreign keys, the information captured in those columns gives the requisite context and meaning to rest of the information in that row.
+Columns that have schema specifications may not have such context and meaning and thus would be confusing to impose at the application end.
 
@@
 As columns from the Movie Information table and the User Information table have been filled with values, there will be a new row added to the Movie Information and User Information table.
-Any issues with adding to the Movie Information table will have to be resolved by the local database administrator.
-If the user decides to try and add this row without a value for the ``Website that user has obtained rating from'' column, a primary key for the User Information table, the application will prompt the user to enter the value prior to allowing a database push.
-We did not feel that this was a strong or particularly confusing change to force the user to make and it would help reduce the number of issues pushed upon the database administrator.
+Any issues with adding to the Movie Information table (e.g., missing a value for a column denoted not null in the database) will have to be resolved by the local database administrator.
+Changes that are definitely going to be rejected (e.g., adding in a row without a value for the primary key(s) will be rejected by the data source without needing input from the database administrator).
 
-\begin{table}[h!]
+\begin{table}[!ht]
 \centering
@@
 
-\begin{table}[h!]
+\begin{table}[!ht]
 \centering
@@
 
-When the user adds a row, the change that is logged to the transformation script has the following structure: addRow(tableName(columnName1:valuesUserPutIn1, columnName2:valuesUserPutIn2, ..., columnNameN:valuesUserPutInN), timestamp).
+When the user adds a row, the change that is logged to the transformation script has the following structure: \\ \\
+addRow(tableName(columnName1:valuesUserPutIn1, \\
+ \indent columnName2:valuesUserPutIn2, \\
+ \indent ..., \\
+ \indent columnNameN:valuesUserPutInN), timestamp).
 
@@
 \label{sec:remove_row}
-When removing a row, the user has to delete all values from columns that belong to a single table before the row can be successfully deleted in the database.
-This design decision was made to avoid users accidentally deleting whole rows when they just want to delete a single value from a column.
-We thought that this reflects how a user would normally use the system \textemdash if he/she really wanted to delete a row, they would go delete the whole row already.
+There are two methods for a user to delete a row in the data store.
+The first method involves selecting the ''Remove Row'' button while the second method is to delete all values from cells associated with that row in the table.
+Imposing only cell-at-a-time deletion when manipulating from the table helps avoid incidents where users accidentally delete whole rows when they just want to delete a single cell.
+We thought that this reflects how a user would normally use the system \textemdash if a user really wanted to delete a row, the whole row would be deleted anyways.
 As we felt that this accurately reflected a normal usage situation, we did not consider this as a constraint that imposed extra annoyance upon the user.
 
-When the user removes a row, the change that is logged to the transformation script has the following structure:
-removeRow(tableName(primaryKey1:\\primaryKeyValue1, primaryKey2:primaryKeyValue2, ..., primaryKeyN:\\primaryKeyValueN), timestamp).
-We are just using the primary key to identify which specific row to delete and have decided to let the database system handle delete propagation (i.e., deleting the rows that other tables have which use values referencing the ones in the deleted row).
+When the user removes a row by the ``Remove Row'' button, the change that is logged to the transformation script has the following structure: \\ \\
+removeRow(tableName(primaryKey1:primaryKeyValue1, \\ 
+\indent primaryKey2:primaryKeyValue2,\\
+\indent ..., \\
+\indent primaryKeyN:primaryKeyValueN), \\
+\indent timestamp). \\ \\
+We are use the primary key to identify which specific row to delete.
+The database system will handle delete propagation (i.e., deleting the rows that other tables have which use values referencing the ones in the deleted row).
+
+Removing values of all cells related to that row may also effectively delete the row.  Further value-editing details are described in Section \ref{sec:edit_value}.
 
@@
 \label{sec:add_col}
-When adding a column, there was a choice between having a column added to all tables that have columns in the current view or having the column added to one of the tables that had columns in the view.
-Originally we had the column added to all tables that have columns in the current view but quickly realized that it would lead to a lot of update problems along with the fact that it would take up extra memory.
+When adding a column, we considered two main possibilities: having a column added to all tables that have projections in the current view or adding to only one table.
+Originally we chose to have the column added to all tables that have representation in the current view but quickly realized that it would lead to a lot of update problems including potentially increasing the number of otherwise null-valued rows.
+There would also be the problem of using extra memory.
 In the end, we decided to have the column added to the table with the most columns in the current view.
-If tables have an equal number of columns in the view, the table can be chosen in one of three ways:
+If tables have an equal number of columns in the view, final destination for the column can be chosen in one of three ways:
 \begin{itemize}
@@
 
-We decided to choose the second method to resolve any possible conflicts between table choice when adding columns.
-The add column change is logged as follows: addCol(tableName, nameOfNewCol, timestamp).
+We decided to choose the second method to resolve any possible conflicts between table choice when adding columns. The add column change is logged as follows: \\ \\
+addCol(tableName, nameOfNewCol, timestamp).
 
@@
 It could be argued that the user should be able to do whatever he/she wishes upon the data as that is the point of enabling users from all backgrounds to use and update data but upon weighing the pros and cons of this issue, it was decided that it was too easy for the user to cause major damage to the data source if primary key removal was allowed.
-If the user wishes to remove a primary key column, he/she could contact the database administrator.
-The remove column change would be logged into the transformation script as follows: removeCol(tableName, nameOfColToRemove, timestamp).
+This would severely limit the availability and use of the data when scaled to allow for concurrent users.
+Thus we decided on a requiring that users who wish to remove a primary key column would require the aide of the database administrator.
+The ''Remove Column'' change would be logged into the transformation script as follows: \\ \\
+removeCol(tableName, nameOfColToRemove, timestamp).
 
@@
 \label{sec:rename_col}
-Although the situation where a user would rename a column is uncommon, we decided to enable support for this use case as it could potentially come up.
-The transformation script syntax for this change is: renameCol(tableName, oldNameofCol, newNameOfCol, timestamp) where the tableName is the name of the table that the column to be renamed is located in, oldNameOfCol is the old name of the column used to identify which column to rename, and newNameOfCol is the new name of the column.
+Although the situation where a user would rename a column is uncommon, we decided to enable support for this use case regardless.
+The transformation script syntax for this change is: \\ \\ renameCol(tableName, oldNameofCol, newNameOfCol, timestamp).
 
@@
 \label{sec:edit_value}
-When editing a value, a user can make either a 1-1 or 1-0 change (i.e., a 1-1 change is when the user can change one value to another while a 1-0 change would be when a user deletes the value).
-As with the add row functionality (see Section \ref{sec:add_row}), if a user ends up violating schema constraints when editing a value, the database administrator will have to approve/deal with the changes.
-Some of the possible schema violations include changing a unique value to a non-unique one when the column specifies that it needs to have unique values or deleting a value from a column that has specified that it cannot be null.
-The edit a value change would be logged into the transformation script as follows: editValue(tableName(columnNameOfEditedValue1: editedValue1, columnNameOfEditedValue2:editedValue2, ..., columnNameOfEditedValueN:editedValueN).
+When editing a value, a user can make one of two types of changes, defined as follows:
+\begin{itemize}
+\item 1-1 \textemdash change from one value to another
+\item 1-0 \textemdash value is deleted
+\end{itemize}
+
+As with the add row functionality (see Section \ref{sec:add_row}), if a user ends up violating schema constraints when editing a value, the database administrator will have to handle the changes.
+Some of the possible schema violations include changing a unique value to a non-unique one when the column specifies that it needs to have unique values or deleting a value from a column that has forbidden null values.
+The ''Edit Value'' change would be logged into the transformation script as follows: \\ \\
+
+\noindent editValue(tableName(originalPrimaryKey1,.., originalPrimaryKeyN, \\
+\indent columnNameOfEditedValue1: editedValue1, \\
+\indent columnNameOfEditedValue2:editedValue2, \\
+\indent ..., \\
+\indent columnNameOfEditedValueN:editedValueN). \\ \\
+
+The originalPrimaryKey1 to originalPrimaryKeyN variables hold the values of the primary keys for the row prior to any changes.
+The initial primary key values are needed to identify which row is to be changed; if one of the primary key values have been edited, we would not be able to correctly identify the rows to be changed without the initial primary key values.
 
@@
 In the end, we decided on a mixed-initiative approach where the user would push if certain types or a certain number of changes have occurred while also allowing the user to choose when to push changes.
-The primary concern was that schema changes (see Section \ref{sec:add_row}, \ref{sec:remove_row}, \ref{sec:add_col}, and \ref{sec:remove_col}) were important enough to push right away as the longer the wait period, the potential for concurrent users becomes greater.
-Value changes are considered less important; although value changes from one user may conflict with another user, it was decided to use the timestamp to resolve conflicts.
-The application can push to the remote data source after a certain number of value changes.
-As for quantifying the number of changes that should occur before the application automatically pushes to the data source, that has yet to be determined.
+The primary concern was that schema changes were important enough to push right away (see Section \ref{sec:add_row}, \ref{sec:remove_row}, \ref{sec:add_col}, and \ref{sec:remove_col}) especially since we acknowledged that as the wait period grew, the potential effects of bad information grew accordingly.
+If this project is later extended to include multiple concurrent users, the longer schema changes are withheld, the greater the potential is for affecting a large number of users.
+Value changes are considered less important; for value changes that conflict from user to user, we elected to refer to the timestamp for resolution decision.
+The application can push to the remote data source after a threshold number of changes.
+We have left the definition of the automatic push threshold as a future work.
 
 \subsection{Proof of Concept}
-After designing the transformation language, we decided to implement the scenario in order to show that our transformation language and design decisions could work to update a data source.
-The only extra step that was not included in \cite{arniThesis} was to allow for the user to choose which changes to sync to the remote data source.
-We created an interface using Netbeans and \textcolor{red}{add screen shots!!
-Kailang can write about all the cool things she did here!!
-:)}
+After designing the transformation language, we decided to implement the scenario from \cite{arniThesis} in order to show that our transformation language and design decisions could work to update a data source.
+A critical feature that we added was to allow for the user to choose which changes to sync to the remote data source.
+We created an interface using Netbeans that enabled users to load data from data source, make changes (while selecting whether to sync those changes), and push selected changes to remote data source.
 
+\figref{fig:load} shows the original MOVIESTAR table data we extracted and loaded from the data source.
+We are only showing three columns: the first is a unique id of a movie star (i.e.
+the primary key of the table), the second is the name of the movie star, and the third is the gender of the actor.
+Currently, we only load the specific data and columns, but this can be easily extended to letting users choose which columns and rows to fetch.
+Every time ``Pull'' is clicked, we discard all the un-pushed changes and load the data from the remote data source.
+Keeping the remote data source distinct from the application view means that a ''Pull'' without a ''Push'' is effectively a local ''Undo''.
+
+\begin{figure} [!ht]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{load}
+  \caption{Original data loaded from remote data source}
+  \label{fig:load}
+\end{figure}
+
+Now, we can try to make some changes.
+We will track all ``Map'' changes (i.e., adding/deleting a row, and editing the value) while ``Reshape'' changes (i.e.,  creating/delete/renaming a column) are only logged to the transformation script if the user wishes for the changes to be synced to the remote data source.
+\figref{fig:unsync} shows a few changes (add column ``height'' and ``weight'', and rename the first column to be ``ID'') we made to the original data that we chose to not sync and keep locally.
+\figref{fig:sync} shows some changes (add column ``age'', add a row, and update some cells) that we chose to log and sync.
+\\
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\columnwidth]{unsync}
+  \caption{Changes that users choose not to sync}
+  \label{fig:unsync}
+\end{figure}
+
+\begin{figure} [!ht]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{sync}
+  \caption{Changes that will be logged and synced}
+  \label{fig:sync}
+\end{figure}
+\newpage
+Synced changes will not apply to the remote data source immediately.
+Instead, we generate a transformation script, and log every synced change to it using our transformation language.
+For example, operations in \figref{fig:sync} will be logged in our script and once the user selects the ``Push'' button, we will translate and and apply the transformation script to the remote data source.
+If we then click on ``Load'', the tool will load data in from the data source again (see \figref{fig:push}).
+We can see that the data loaded in from the data source is slightly different from the data we saw before reloading.
+This is because only part of the changes made to our local data will be synced to the remote data source.
+If we load again, some changes we made to the local data copy will disappear.
+%\newpage
+\begin{figure} [!ht]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{push}
+  \caption{Modified data re-load from remote data source after push}
+  \label{fig:push}
+\end{figure}
+%\clearpage
+
+And also, the execution result of our transformation script and our translation of these scripts are proved to correct by comparing these results to the data set before we ``Push'' and re-``Load''.
+
+%\newpage
 \section{Related Work}
@@
 \subsection{View Maintenance}
-As our application takes a mixed-initiative approach to updating the remote data source, we also thought that we might be able to get some ideas on how often to push any possible schema mapping/schema changes as it seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.
-Through looking at how views and indices are maintained in a large distributed database \cite{agrawal1997efficient}, we realized that many of the design decisions we made for our transformation language (see Section \ref{sec:transformation_language}) fit quite well with the model suggested in the paper.
-For example, their discussion on the different ways to push updates and the decisions made on whether the system or the application should handle certain update issues do mirror our decisions.
+Due to our use of a remote data source that doesn't know about update details made by the application, we began our view maintenance research in data warehousing environments.
+We considered the \textit{Eager Compensating Algorithm} (ECA) \cite{zhuge1995view} which described triggering events to update the warehouse or the application's materialized view.
+While this approach may be helpful in correcting for inconsistencies arising from decoupling the data source and application, it requires quite a bit of overhead to regularly connect and creates the opportunity for anomalies (consistency issues that arise when updating views while base data is being changed) \cite{zhuge1995view}.
+
+We implemented an application that takes a mixed-initiative approach to updating the remote data source making it important for us to be aware of how often to push any possible schema mapping/schema changes.
+This seems to be a comparable problem to how materialized views need to push its changes to a remote database \cite{agrawal1997efficient, agrawal2009asynchronous, zhou2007lazy}.
+Through looking at how views and indices are maintained in a large distributed database \cite{agrawal1997efficient}, we noted that the model suggested in the papers validated many of the design decisions we made for our transformation language (see Section \ref{sec:transformation_language}).
+For example, the discussion on the different ways to push updates and how to handle update discrepancies (source versus the application) \cite{agrawal1997efficient} is reflected in our implementation.
 
@@
 
-Another extension of the project could be investigating how the data transformations could be integrated into data provenance to help with data conflict resolutions \cite{arniThesis}.
+Another extension of the project could be investigating how the data transformations could be integrated into data provenance to help with automatic data conflict resolutions \cite{arniThesis}.
 As the project currently stands, the remote application can push any type of data to the data source.

### 2015-04-15 src/mainGUI.java
Delete mainGUI.java
Remove the old GUI

@@ -1,122 +1,0 @@
-// taken from https://docs.oracle.com/javase/tutorial/displayCode.html?code=https://docs.oracle.com/javase/tutorial/uiswing/examples/layout/GridBagLayoutDemoProject/src/layout/GridBagLayoutDemo.java
-import java.awt.*;
-
-import javax.swing.JButton;
-import javax.swing.JFrame;
-import javax.swing.JTextField;
- 
-public class mainGUI {
-    final static boolean shouldFill = true;
-    final static boolean shouldWeightX = true;
-    final static boolean RIGHT_TO_LEFT = false;
- 
-    public static void addComponentsToPane(Container pane) {
-        if (RIGHT_TO_LEFT) {
-            pane.setComponentOrientation(ComponentOrientation.RIGHT_TO_LEFT);
-        }
- 
-        JButton button;
-    pane.setLayout(new GridBagLayout());
-    GridBagConstraints c = new GridBagConstraints();
-    if (shouldFill) {
-    //natural height, maximum width
-    c.fill = GridBagConstraints.HORIZONTAL;
-    }
- 
-    button = new JButton("Add Row");
-    if (shouldWeightX) {
-    c.weightx = 0.5;
-    }
-    c.fill = GridBagConstraints.HORIZONTAL;
-    c.gridx = 0;
-    c.gridy = 0;
-    pane.add(button, c);
-    
-    button = new JButton("Delete Row");
-    if (shouldWeightX) {
-    c.weightx = 0.5;
-    }
-    c.fill = GridBagConstraints.HORIZONTAL;
-    c.gridx = 0;
-    c.gridy = 1;
-    pane.add(button, c);
- 
-    button = new JButton("Add Column");
-    c.fill = GridBagConstraints.HORIZONTAL;
-    c.weightx = 0.5;
-    c.gridx = 1;
-    c.gridy = 0;
-    pane.add(button, c);
- 
-    button = new JButton("Delete Column");
-    c.fill = GridBagConstraints.HORIZONTAL;
-    c.weightx = 0.5;
-    c.gridx = 1;
-    c.gridy = 1;
-    pane.add(button, c);
-    
-    button = new JButton("Edit Cell Value");
-    c.fill = GridBagConstraints.HORIZONTAL;
-    c.weightx = 0.5;
-    c.gridx = 2;
-    c.gridy = 0;
-    pane.add(button, c);
- 
-    JTextField textField = new JTextField(25);
[…]
